---
title: 'JHU Reproducible Research: Project 1'
author: "Shams Shahadat"
date: "July 17, 2017"
output:
  md_document: default
  html_document: default
---

```{r dir, echo=FALSE, results='hide'}
# This checks the Working directory
getwd()
```

```{r clear_variables,echo=FALSE, results='hide'}
# This clears/removes all variables from the environment
rm(list=ls(all=TRUE))
```

## Introduction

A description of the project/assignment requirements is presented on repo file **README.md** (Forked Content from [Roger D. Peng @ GitHub](https://github.com/rdpeng)).

## Methods/Results

### Loading and Preprocessing the Data

```{r load_data,echo=TRUE, results='hide'}
# Load data file - activity.csv
activity <- read.csv("activity.csv", sep = ",")
# head(activity)
summary(activity)
```

### The Mean Total Number of Steps per Day

The number of steps were grouped by date and intervals using **reshape2** package. Days with no steps were removed.

```{r mean_steps_per_day,echo=TRUE, results='hide'}
library(reshape2)
activity_melt <- melt(activity[which(!is.na(activity$steps)), ], id.vars = c("date", "interval"))
head(activity_melt)
steps_sum <- dcast(activity_melt, date ~ variable, sum)
head(steps_sum)
```
Then the summary can present the mean and median data.

```{r mean, echo=TRUE, results='hide'}
summary(steps_sum$steps)
mean_steps <- mean(steps_sum$steps)
median_steps <- median(steps_sum$steps)
```

Then the histogram of the total number of steps taken each day can displayed.

```{r total_steps_histogram,echo=TRUE,results='asis'}
hist(steps_sum$steps, main = "Total Steps per Day Histogram",
  xlab = "Total Steps per day", ylab = "Number of Days", breaks = 10)
abline(v = mean(steps_sum$steps), lty = 1, lwd = 2, col = "red")
abline(v = median(steps_sum$steps), lty = 2, lwd = 2, col = "steel blue")
legend(x = "topright", c("Mean", "Median"), col = c("red", "steel blue"),lty = c(1, 2), lwd = c(2, 2))
```

The mean and median are respectively `r mean_steps` and `r median_steps`, about the same.

### The Average Daily Activity Pattern

A time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken averaged across all days was constructed.

```{r mean_interval,echo=TRUE,results='asis'} 
stepsmeaninterval <- dcast(activity_melt, interval ~ variable, mean, na.rm = TRUE)
plot(stepsmeaninterval$interval, stepsmeaninterval$steps, ty = "l", 
     xlab = "5-Minute Interval", ylab = "Average Number of Steps", main = "Average Number of 
     Steps Over All Days vs Time interval")
```

The maximum average can be calculated.

```{r moststeps_interval,echo=TRUE,results='hide'}
maxsteps_interval <- stepsmeaninterval$interval[which.max(stepsmeaninterval$steps)]
maxsteps_interval
```
As seen in the plot, `r maxsteps_interval`th interval, on average across all the days, contains maximum number of steps.

### Imputing Missing Values

```{r ns_data_effect,echo=TRUE,results='hide'}
na_data_total <- sum(is.na(activity$steps))
# Percentage of rows with missing values
effect <- sum(is.na(activity$steps))*100/nrow(activity) 
```
There are `r na_data_total` missing values in the dataset (i.e. the total number of rows with NAs). In order to impute missing values, finding intervals with the missing values should be identified. In other words, find out whether there are days with all time intervals reporting NA step values. The missing data for a day can be replaced by the time average over all other days. A new dataset is created.

```{r new_activity,echo=TRUE,results='asis'}
new_activity <- split(activity, activity$interval)
new_activity <- lapply(new_activity, function(x) {
  x$steps[which(is.na(x$steps))] <- mean(x$steps, na.rm = TRUE)
  return(x)
})
new_activity <- do.call("rbind", new_activity)
row.names(new_activity) <- NULL
new_activity <- split(new_activity, new_activity$date)
df <- lapply(new_activity, function(x) {
  x$steps[which(is.na(x$steps))] <- mean(x$steps, na.rm = TRUE)
  return(x)
})
new_activity <- do.call("rbind", new_activity)
row.names(new_activity) <- NULL
#head(new_activity)
```
The time intervals form a disjoint partitioning of 24 hrs, i.e. 1 day is found to be erroneous meaning the time interval for each day corresponds to about 40 hours. This refutes the intervals being disjoint.

```{r new_activity_histogram,echo=TRUE,results='asis'} 
library(reshape2)
activity_melt2 <- melt(new_activity, id.vars = c("date", "interval"))
new_steps_sum <- dcast(activity_melt2, date ~ variable, sum, na.rm = TRUE)
#head(new_steps_sum)
#summary(new_steps_sum$steps)
new_mean_steps <- mean(new_steps_sum$steps)
new_median_steps <- median(new_steps_sum$steps)
hist(new_steps_sum$steps, main = "Total Steps per Day Histogram with Imputed Missing Data", 
  xlab = "Total Steps per day", ylab = "Number of Days",breaks = 10)
abline(v = mean(steps_sum$steps), lty = 1, lwd = 2, col = "red")
abline(v = median(steps_sum$steps), lty = 2, lwd = 2, col = "black")
legend(x = "topright", c("Mean", "Median"), col = c("red", "black"), lty = c(2, 1), lwd = c(2, 2))
```

The new mean and median are respectively `r new_mean_steps` and `r new_median_steps`. As expected the mean is almost the same as the first part; however, the old median slightly changes to the new mean. This is due to the fact as the missing data constitue `r effect`% of the total number (row) data.

### Differences in Patterns between Weekdays and Weekends

A new column is created describing if the date is a weekday or weekend.

```{r day_type, echo=TRUE,results='hide'}
library(lubridate)
weekends <- which(weekdays(as.Date(new_activity$date)) == "Saturday"
|weekdays(as.Date(new_activity$date))=="Sunday")
weekdays <- which(weekdays(as.Date(new_activity$date)) != "Saturday" 
& weekdays(as.Date(new_activity$date)) != "Sunday")
temp <- c(rep("a", length(new_activity)))
temp[weekends] <- "weekend"
temp[weekdays] <- "weekday"
length(temp)
names(temp) <- "day"
new_activity <- cbind(new_activity, temp)
names(new_activity)[4] <- "day"
```

Steps taken over each interval was averaged across weekdays and weekends.

```{r}
new_activity_split <- split(new_activity, new_activity$day)
stepsmean_interval <- lapply(new_activity_split, function(x) {
  temp <- aggregate(x$steps, list(x$interval), mean)
  names(temp) <- c("interval", "steps")
  return(temp)
})
```

```{r, echo=TRUE, results='hide'}
## stepsmean_interval
stepsmean_interval <- do.call("rbind", stepsmean_interval)
weekdays <- grep("weekday" ,row.names(stepsmean_interval))
weekends <- grep("weekend" ,row.names(stepsmean_interval))
temp <- c(rep("a", length(stepsmean_interval$steps)))
temp[weekdays] <- "weekdays"
temp[weekends] <- "weekends"
stepsmean_interval <- cbind(stepsmean_interval, temp)
row.names(stepsmean_interval) <- NULL
names(stepsmean_interval)[3] <- "day"
head(stepsmean_interval)
tail(stepsmean_interval)
```

```{r}
library(ggplot2)
ggplot(stepsmean_interval, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) 
```
The mean number of steps taken over the weekdays and weekends is calculated using the following:

```{r}
stepsdatamelt <- melt(stepsmean_interval, id.vars = c("interval","day"))
dcast(stepsdatamelt, day ~ variable, mean) # Average steps
```

## Conclusion

As a whole, the project assignement was an excellent introduction to using **knitr** and **R markdwon**. In terms of data analysis, the answer to the questions still need to be justified to validate the reproducibility process.
